Lossy compression of DNA sequencing quality data

### 2015-04-24 manuscript/q-compression.tex
Rewrote some of the preprocessing section.

We evaluate effects of information loss from quality value compression
on quality control steps of read filtering and trimming, which were
performed using Sickle~\citep{sickle}, and make comparison to
uncompressed data.

@@

The majority of compression methods retain more base-pairs after
preprocessing than the uncompressed sequences (Figure
\ref{fig:preprocessing}). In general, as a given compression model
increases in complexity, i.e., as the number of profiles, polynomial
degrees, or rate increases, the amount of base-pairs kept approaches
the number base-pairs kept using the uncompressed sequences. The
compression methods on the \textit{Mus musculus} data set have the
greatest proportion of retained base-pairs compared to the
uncompressed sequences. The \textit{Escherichia coli} str. K-12 MG1655
MiSeq data set has the smallest range.

The 2-bin approach is the only compression method that results in a
higher number of filtered base-pairs across all data sets. Sickle uses
a sliding window approach to smooth the read quality values before it
trims the sequence based on a specific quality threshold. In the 2-bin
approach, there is not an even distribution of values per bin.  In
other words, \emph{bad} quality values may range from 0-33, whereas
\emph{good} values may only range from 34-40. Thus, mid-range quality
values that are above the threshold (20 by default) are set below the
quality threshold when compressed, resulting in an increased number of
filtered bases.

The 0-degree polynomial regression results in the highest proportion
of bases kept. If the mean quality value of the read is above the
filtering threshold, then no base-pairs are trimmed. Only reads that
are comprised of mainly low quality values will be filtered.

It is important to highlight that the even though a compression method
may result in the same number of filtered base calls as the
uncompressed sequences, it does not mean the \emph{same} base calls
were filtered. The 1st-degree and 5th-degree polynomial regression of
the \textit{Rhodobacter sphaeroides} fragment library filters nearly
as many bases as each other. However, if we examine the specific reads
filtered, the 5th-degree polynomial regression discards approximately
two thirds less reads than the 1st-degree polynomial regression that
are kept by the uncompressed reads (4,640 and 12,066 reads,
respectively) (Supplemental Table 2).

>>>>

We evaluate effects of information loss from quality value compression
on quality control steps of trimming and read filtering, which were
performed using Sickle~\citep{sickle}, and make comparison to
uncompressed data.
Sickle starts at the ends of the sequence and uses a sliding window ($0.1 \times$ the read length) to find locations where the window's average quality falls below a given threshold (20 by default).
The sequence is then trimmed from this position to the respective end.
%uses a variable sliding window approach where base qualities are averaged over a window ($0.1 \times$ the read length) and trims from the position where the window's average quality falls below a given threshold (20 by default).
If the trimmed sequence's length is less than a certain threshold (20 by default), then the sequence is discarded.

@@

Preprocessing involves two steps: \emph{trimming} the poor quality regions of the sequences and \emph{discarding} sequences that are deemed to be poor quality overall.
After trimming, the majority of compression methods retain more base-pairs than the uncompressed sequences (Figure
\ref{fig:preprocessing}). In general, as a given compression model
increases in complexity, i.e., as the number of profiles, polynomial
degrees, or rate increases, the amount of base-pairs kept approaches
the number base-pairs kept using the uncompressed sequences. The
compression methods on the \textit{Mus musculus} data set have the
greatest proportion of retained base-pairs compared to the
uncompressed sequences. The \textit{Escherichia coli} str. K-12 MG1655
MiSeq data set has the smallest range.

The 2-bin approach is the only compression method that results in a
higher number of trimmed base-pairs across all data sets. Sickle uses
a sliding window approach to smooth the read quality values before it
trims. In the 2-bin
approach, there is not an even distribution of values per bin.  In
other words, \emph{bad} quality values may range from 0-33, whereas
\emph{good} values may only range from 34-40. Thus, mid-range quality
values that are above the threshold (20 by default) are set below the
quality threshold when compressed, resulting in an increased number of
trimmed bases.

The 0-degree polynomial regression results in the highest proportion
of bases kept. If the mean quality value of the read is above the
filtering threshold, then no base-pairs are trimmed. Only reads that
are comprised of mainly low quality values will be discarded.

It is important to highlight that even though a compression method
may result in the same number of trimmed bases as the
uncompressed sequences, it does not mean the \emph{same} bases
were trimmed. The 1st-degree and 5th-degree polynomial regression of
the \textit{Rhodobacter sphaeroides} fragment library filter nearly
as many bases as each other. However, if we examine the specific reads
discarded, the 5th-degree polynomial regression discards approximately
two thirds less reads than the 1st-degree polynomial regression that
are kept by the uncompressed reads (4,640 and 12,066 reads,
respectively) (Supplemental Table 2).
